Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-10-14 demystify.tex
update

International students, especially those from less well-known universities and smaller countries, often wonder if they should apply to a CS PhD program in the US. Specifically, they are worried about (i) the \textbf{difficulty of getting admitted} and (ii) the \textbf{cost of graduate study} in the US. This section addresses these concerns.

First, applying to a good US university \emph{should not} be any harder than at schools in other countries. In fact, it might even be more flexible since CS PhD in the US \emph{do not} require having an MS or a research topic, proposal, or adviser in advance (\autoref{sec:non-us-differences} compares CS PhD study in the US to other countries). If you believe you have a chance in other countries, e.g., South Korea, Singapore, Germany, UK, Japan, and Australia, then you will surely have a chance in the US as well.

Second, PhD students in CS \emph{do not} need to worry about funding, especially at good R1
universities in the US. If you are admitted, you will almost certainly
receive \emph{full funding} (\autoref{sec:funding}) to support your study. This includes tuition,
health insurance, and stipend (i.e., you get paid for your study). Moreover, depending on the university,
you may receive additional benefits such as summer pay, laptops, and traveling to conferences and workshops.

The rest of this document will provide you with more details on the CS PhD admission process in the US, including how to prepare your application, what to expect during process, what to do with acceptance (and how to deal with rejection), and so on. The goal is to help you understand the process and make the best decisions for your future.

@@

@@

Typically, it takes 5--7 years for CS PhD in the US.  This can be longer than CS PhDs in other countries (\autoref{sec:msrequirement}), which might require MS first. Within these 5--7 years, CS PhD students often take a ``leave of absence'' for 1--2 semesters to do internships at companies and research labs.

@@

The first two years you typically take coursework (somewhat equivalent to an MS study), find an adviser, and learn how to do research.  The next 2--3 years you focus on your research, form a dissertation topic, and get results published. The last 1--2 years you continue to publish, write and defend your dissertation, and look for a job.
In many cases you might take a summer or two off to do an internship to get additional research opportunities.
The PhDComics figure above shows the ``ambition'' level of a PhD student over their years of study (they miss the 6--7th year when the ambition is \emph{``Just let me graduate''}).

% \begin{commentbox}[Vu:]
%     I start my PhD with an MS, and it took me 7 years (Fall'2007-- Fall'14). I spent half a year doing an internship at the Naval Research Lab.
%   \end{commentbox}

>>>>

International students, especially those from less well-known universities and smaller countries, often wonder if they should apply to a CS PhD program in the US. Specifically, they are worried about (i) the \textbf{difficulty of getting admitted} and (ii) the \textbf{cost of graduate study} in the US. Part of the reasons is the lack of information and guidance on the admission process in the US (let's face it, public forums like Facebook communities and Reddit are full of confusion and concerns from students).  This section and the rest of this document aim to address these concerns.

First, applying to a good US university \emph{should not} be any harder than at schools in other countries. It might even be more flexible since CS PhD in the US \emph{do not} require having an MS or a research topic, proposal, or adviser in advance (\autoref{sec:non-us-differences} compares CS PhD study in the US to other countries). 
If you believe you have a chance in other countries, e.g., South Korea, Singapore, Germany, UK, Japan, and Australia, then you will surely have a chance in the US as well.

Second, PhD students in CS \emph{do not} need to worry about funding, especially at good R1
universities in the US. 
If you are admitted, you will almost certainly
receive \emph{full funding} (\autoref{sec:funding}) to support your study. This includes tuition,
health insurance, and stipend (i.e., you get paid for your study). 
In fact, full funding for CS PhD students is the norm in the US, and I'd go as far as to say that if you are not admitted with full funding, you should be cautious about the program and might want to not accept the offer.
Moreover, depending on the university, you may receive additional benefits such as summer pay, laptops, and traveling to conferences and workshops.

@@

Another goal is to explain the \emph{``why''} behind the admission process, e.g., why LoRs are important, why you should not draft your own LoR, why you should contact professors, etc.

@@

Typically, it takes 5--7 years for CS PhD in the US.  This can be longer than CS PhDs in other countries (\autoref{sec:msrequirement}), which might require MS first.

@@

The first two years you typically take coursework (somewhat equivalent to an MS study), find an adviser, and learn how to do research.  The next 2--3 years you focus on your research, form a dissertation topic, and get results published. The last 1--2 years you continue to publish, write and defend your dissertation, and look for a job.
Within these 5--7 years, CS PhD students often take a ``leave of absence'' for 1--2 semesters or summer to do internships at companies and research labs.

The PhDComics figure above shows the ``ambition'' level of a PhD student over their years of study (they miss the 6--7th year when the ambition is \emph{``Just let me graduate''}).

\begin{commentbox}[Vu:]
    I start my PhD with an MS, and it took me 7 years (Fall'07-- Fall'14). I spent half a year doing an internship at the Naval Research Lab. My PhD did take a bit longer than usual, but allows me to explore various research areas and topics.
  \end{commentbox}
